## Effects of non-standard roles


Recipes steps can keep column(s) of your data set that should not be treated as outcomes or predictors. For example, a unique identifier column or some other ancillary data that can be used to troubleshoot issues during model development. 

As an example, you might have a date column that is transformed into features using [step_date()] and.or [step_holiday()]. 

```{r, results = "hide", message = FALSE, warning = FALSE}
  library(tidymodels)
  tidymodels_prefer()

  data(Chicago, package = "modeldata")
  chi_rec <- recipe(ridership ~ ., data = Chicago) %>% 
    # Since `date` is symbolically on the right-hand side of the formula, it
    # is treated as a predictor (i.e., has a role of "predictor")
    step_holiday(date) %>% 
    step_date(date)
```

At this pont, `date` is kept in the dataset as-is, and will be treated as a predictor (and will probably be converted to an integer column by the modeling function). 

One option is to remove date since it is not directly used as a model term. [step_rm()] is a good way to accomplish this. 

```{r}
  chi_rec_rm <- chi_rec %>% step_rm(date)
```

Another approach, especially helpful during model development, is to change the role of the column to some other value via [update_role()]. 

```{r}
  chi_rec_role <- chi_rec %>% update_role(date, new_role = "date")
```

In this case, the column is still available in the data and will be carried through the analysis (e.g. during resampling). This can be helpful since, it there is a poor prediction, this column may be helpful in diagnosing the problem. 

The downside to chaining the role of the column is that it is needed at prediction-time. This can impact development tools (such as tuning functions) and would also be required to be in the data being predicted once the model is deployed. 

If using a non-standard role for a column, it highly recommended that you use the recipe within a workflow and specify `bake_dependent_roles` as described in [hardhat::default_recipe_blueprint()]. This will allow tuning and prediction functions to run without error. 

With the example above:

```{r, warning = FALSE}
  library(hardhat) # require hardhat version >= 1.0.0
  # Make a new blueprint:
  date_bp <- default_recipe_blueprint(bake_dependent_roles = "date")

  chi_wflow <- 
    workflow() %>% 
    add_model(linear_reg()) %>% 
    add_recipe(chi_rec_role, blueprint = date_bp) 

  chi_fit <- fit(chi_wflow, Chicago)
  # This line will fail without the added blueprint:
  predict(chi_fit, head(Chicago))
```
